\textbf{A framework for effective application of machine learning to a microbiome-based classification problem}

### 2020-01-28 submission/manuscript.tex
Get to render on my laptop

\documentclass[11pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={Effective application of machine learning to a microbiome-based classification problem},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

%%% Use protect on footnotes to avoid problems with footnotes in titles
\let\rmarkdownfootnote\footnote%
\def\footnote{\protect\rmarkdownfootnote}

%%% Change title format to be more compact
\usepackage{titling}

% Create subtitle command for use in maketitle
\newcommand{\subtitle}[1]{
  \posttitle{
    \begin{center}\large#1\end{center}
    }
}

\setlength{\droptitle}{-2em}

  \title{\textbf{Effective application of machine learning to a microbiome-based
classification problem}}
    \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
    \author{}
    \preauthor{}\postauthor{}
    \date{}
    \predate{}\postdate{}

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage[table]{xcolor}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{caption}
\usepackage{hyperref}
\usepackage{helvet} % Helvetica font
\renewcommand*\familydefault{\sfdefault} % Use the sans serif version of the font
\usepackage[T1]{fontenc}
\usepackage[labelfont=bf]{caption}

@@

@@

\linenumbers

\subsection{Abstract}\label{abstract}

@@

\subsection{Importance}\label{importance}

@@

\subsection{Background}\label{background}

As the number of people represented in human microbiome datasets grow,
there is an increasing desire to use microbiome data to diagnose
disease. However, the structure of the human microbiome is remarkably
variable among individuals to the point where it is often difficult to
identify the bacterial populations that are associated with diseases
using traditional statistical models. This variation is likely due to
the ability of many bacterial populations to fill the same niche such
that different populations cause the same disease in different
individuals. Furthermore, a growing number of studies have shown that it
is rare for a single bacterial species to be associated with a disease.
Instead, subsets of the microbiome account for differences in health.
Traditional statistical approaches do not adequately account for the
variation in the human microbiome and typically consider the protective
or risk effects of each bacterial population separately (1). Recently,
machine learning (ML) models have grown in popularity among microbiome
researchers as our ability to sample large numbers of individuals has
grown; such models can effectively account for the interpersonal
microbiome variation and the ecology of disease.

@@

In this study, we provided steps toward standardization of machine
learning methods for microbiome studies which are often poorly
documented and executed. To showcase a rigorous ML pipeline and to shed
light on how ML model selection can affect modeling results, we
performed an empirical analysis comparing the predictive performance,
interpretability and training times of seven modeling approaches with
the same dataset and pipeline. We built three linear models with
different forms of regularization: L2-regularized logistic regression
and L1 and L2-regularized support vector machines (SVM) with a linear
kernel. We also trained four non-linear models: SVM with radial basis
function kernel, a decision tree, random forest, and gradient boosted
trees. We compared their predictive performance, interpretability, and
training time. To demonstrate the performance of these modeling
approaches and our pipeline on a case study, we used a data from a
previously published study that sought to classify individuals as having
healthy colons or colonic lesions based on the 16S rRNA gene sequences
collected from fecal samples (4). This dataset was selected because it
is a relatively large collection of individuals (N=490) connected to a
clinically significant disease where there is ample evidence that the
disease is driven by variation in the microbiome (2, 4, 5, 32). With
this dataset, we developed an ML pipeline that can be used in many
different scenarios for training and evaluating models. This framework
can be easily applied to other host-associated and environmental
microbiome datasets. We also provided an aspirational rubric for
evaluating the rigor of ML practices applied to microbiome data
{[}Figure S1{]} to urge the audience to be diligent with with their
study design, model selection, development, evaluation and
interpretation.

\subsection{Results}\label{results}

@@

Model selection requires tuning hyperparameters. Hyperparameters are
parameters that need to be specified or tuned by the user, in order to
train a model for a specific modeling problem. For example, when using
regularization, C is a hyperparameter that indicates the penalty for
overfitting. Hyperparameters are tuned using the training data to find
the best model. We selected hyperparameters by performing repeated
five-fold cross-validation (CV) on the training set {[}Figure 1{]}. The
five-fold CV was also stratified to maintain the overall case and
control distribution. We chose the hyperparameter values that led to the
best average CV predictive performance using the area under the receiver
operating characteristic curve (AUROC) {[}Figure S1 and S2{]}. The AUROC
ranges from 0, where the model's predictions are perfectly incorrect, to
1.0, where the model perfectly distinguishes between cases and controls.
An AUROC value of 0.5 indicates that the model's predictions are no
different than random. To select hyperparameters, we performed a grid
search for hyperparameter settings when training the models. Default
hyperparameter settings in developed ML packages available in R, Python,
and MATLAB programming languages may be inadequate for effective
application of classification algorithms and need to be optimized for
each new ML task. For example, L1-regularized SVM with linear kernel
showed large variability between different regularization strengths (C)
and benefited from tuning as the default C parameter was 1 {[}Figure
S1{]}.

@@

\textbf{Predictive performance and generalizability of the seven
models.} We evaluated the predictive performance of the seven models to
classify individuals as having healthy colons or SRNs {[}Figure 2{]}.
The predictive performance of random forest model was higher than other
ML models with a median 0.695 {[}IQR 0.650-0.739{]}, though not
significantly (p=0.5; The p-value was manually calculated using the
sampling distribution of the test statistic under the null hypothesis)
(Figure S3). Similarly, L2-regularized logistic regression, XGBoost,
L2-regularized SVM with linear and radial basis function kernel AUROC
values were not significantly different from one another and had median
AUROC values of 0.680 {[}IQR 0.639-0.750{]}, 0.679 {[}IQR
0.643-0.746{]}, 0.678 {[}IQR 0.639-0.750{]} and 0.668 {[}IQR
0.639-0.750{]}, respectively. L1-regularized SVM with linear kernel and
decision tree had significantly lower AUROC values than the other ML
models with median AUROC of 0.650 {[}IQR 0.629-0.760{]} and 0.601 {[}IQR
0.636-0.753{]}, respectively {[}Figure 2{]}. Interestingly, these
results demonstrate that the most complex model (XGBoost) did not have
the best performance and that the most interpretable models
(L2-regularized logistic regression and L2-regularized SVM with linear
kernel) performed nearly as well as non-linear models.

@@

To show the effect of sample size on model generalizability, we compared
cross-validation AUROC values of L2-regularized logistic regression and
random forest models when we subsetted our original study design with
490 subjects to 15, 30, 60, 120, and 245 subjects {[}Figure S4{]}. The
variation in cross-validation performance within both models at smaller
sample sizes was larger than when the full collection of samples was
used to train and validate the models. Because of the high
dimensionality of the microbiome data (6920 OTUs), large sample sizes
can lead to better models.

\textbf{Interpretation of each ML model.} To evaluate the overall
structure of the CRC dataset and show that the structure of the human
microbiome is remarkably variable among individuals with SRNs, we
analyzed the Bray-Curtis distances with an NMDS ordination {[}Figure
S5{]}. The lack of seperation between healthy patients and patients with
SRNs, show that traditional statistical approaches do not help us to
adequately address disease ecology. This is why we often use ML models
not just to predict a health outcome, but also to identify potential
biomarkers for disease. Therefore, model interpretation becomes crucial
for microbiome studies. Interpretability is related to the degree to
which humans can understand the reasons behind a model prediction
(33--35). ML models often decrease in interpretability as they increase
in complexity. In this study, we used two methods to help interpret our
models.

@@

To further highlight the differences between the two interpretation
methods, we used permutation importance to interpret the linear models
{[}Figure S7{]}. When we analyzed the L1-regularized SVM with linear
kernel model using feature rankings based on weights {[}Figure 3{]} and
permutation importance {[}Figure S6{]}, 17 of the 20 top OTUs (e.g., OTU
609, 822, 1239) were deemed important by both interpretation methods.
Similarly, for the L2-regularized SVM and L2-regularized logistic
regression, 9 and 12 OTUs, respectively, were shared among the two
interpretation methods. These results indicate that both methods are
consistent in selecting the most important OTUs.

@@

\subsection{Discussion}\label{discussion}

@@

\subsection{Materials and Methods}\label{materials-and-methods}

@@

\textbf{16S rRNA gene sequencing data.} Stool samples provided by the
patients were used for 16S rRNA gene sequencing to measure bacterial
population abundances. The sequence data used in our analyses were
originally generated by Baxter et al. (available through NCBI Sequence
Read Archive {[}SRP062005{]}, (5)). The OTU abundance table was
generated by Sze et al (45), who processed the 16S rRNA sequences in
mothur (v1.39.3) using the default quality filtering methods,
identifying and removing chimeric sequences using VSEARCH, and assigning
to OTUs at 97\% similarity using the OptiClust algorithm (43, 46, 47);
(\url{https://github.com/SchlossLab/Sze_CRCMetaAnalysis_mBio_2018/blob/master/data/process/baxter/baxter.0.03.subsample.shared}).
These OTU abundances were the features we used to predict colorectal
health of the patients. There were 6920 OTUs. OTU abundances were
subsampled to the size of the smallest sample and normalized across
samples such that the highest abundance of each OTU would be 1, and the
lowest would be 0.

@@

\textbf{Statistical analysis workflow.} Data summaries, statistical
analysis, and data visualizations were performed using R (v.3.5.0) with
the tidyverse package (v.1.2.1). We compared the performance of the
models pairwise by calculating the difference between AUROC values from
the same data-split (for 100 data-splits). We determined if the models
were significantly different by calculating the empirical p-value (2 x
min(\% of AUROC differences \(\geq\) 0, \% of AUROC differences \(\leq\)
0) for the double tail event (e.g., Figure S3).

@@

\subsection{References}\label{references}

\hypertarget{refs}{}
\hypertarget{ref-segata_metagenomic_2011}{}
1. \textbf{Segata N}, \textbf{Izard J}, \textbf{Waldron L},
\textbf{Gevers D}, \textbf{Miropolsky L}, \textbf{Garrett WS},
\textbf{Huttenhower C}. 2011. Metagenomic biomarker discovery and
explanation. Genome Biol \textbf{12}:R60.
doi:\href{https://doi.org/10.1186/gb-2011-12-6-r60}{10.1186/gb-2011-12-6-r60}.

\hypertarget{ref-zeller_potential_2014}{}
2. \textbf{Zeller G}, \textbf{Tap J}, \textbf{Voigt AY},
\textbf{Sunagawa S}, \textbf{Kultima JR}, \textbf{Costea PI},
\textbf{Amiot A}, \textbf{Böhm J}, \textbf{Brunetti F},
\textbf{Habermann N}, \textbf{Hercog R}, \textbf{Koch M},
\textbf{Luciani A}, \textbf{Mende DR}, \textbf{Schneider MA},
\textbf{Schrotz-King P}, \textbf{Tournigand C}, \textbf{Tran Van Nhieu
J}, \textbf{Yamada T}, \textbf{Zimmermann J}, \textbf{Benes V},
\textbf{Kloor M}, \textbf{Ulrich CM}, \textbf{Knebel Doeberitz M von},
\textbf{Sobhani I}, \textbf{Bork P}. 2014. Potential of fecal microbiota
for early-stage detection of colorectal cancer. Mol Syst Biol
\textbf{10}.
doi:\href{https://doi.org/10.15252/msb.20145645}{10.15252/msb.20145645}.

\hypertarget{ref-zackular_human_2014}{}
3. \textbf{Zackular JP}, \textbf{Rogers MAM}, \textbf{Ruffin MT},
\textbf{Schloss PD}. 2014. The human gut microbiome as a screening tool
for colorectal cancer. Cancer Prev Res \textbf{7}:1112--1121.
doi:\href{https://doi.org/10.1158/1940-6207.CAPR-14-0129}{10.1158/1940-6207.CAPR-14-0129}.

\hypertarget{ref-baxter_dna_2016}{}
4. \textbf{Baxter NT}, \textbf{Koumpouras CC}, \textbf{Rogers MAM},
\textbf{Ruffin MT}, \textbf{Schloss PD}. 2016. DNA from fecal
immunochemical test can replace stool for detection of colonic lesions
using a microbiota-based model. Microbiome \textbf{4}.
doi:\href{https://doi.org/10.1186/s40168-016-0205-y}{10.1186/s40168-016-0205-y}.

\hypertarget{ref-baxter_microbiota-based_2016}{}
5. \textbf{Baxter NT}, \textbf{Ruffin MT}, \textbf{Rogers MAM},
\textbf{Schloss PD}. 2016. Microbiota-based model improves the
sensitivity of fecal immunochemical test for detecting colonic lesions.
Genome Medicine \textbf{8}:37.
doi:\href{https://doi.org/10.1186/s13073-016-0290-3}{10.1186/s13073-016-0290-3}.

\hypertarget{ref-hale_shifts_2017}{}
6. \textbf{Hale VL}, \textbf{Chen J}, \textbf{Johnson S},
\textbf{Harrington SC}, \textbf{Yab TC}, \textbf{Smyrk TC},
\textbf{Nelson H}, \textbf{Boardman LA}, \textbf{Druliner BR},
\textbf{Levin TR}, \textbf{Rex DK}, \textbf{Ahnen DJ}, \textbf{Lance P},
\textbf{Ahlquist DA}, \textbf{Chia N}. 2017. Shifts in the fecal
microbiota associated with adenomatous polyps. Cancer Epidemiol
Biomarkers Prev \textbf{26}:85--94.
doi:\href{https://doi.org/10.1158/1055-9965.EPI-16-0337}{10.1158/1055-9965.EPI-16-0337}.

\hypertarget{ref-pasolli_machine_2016}{}
7. \textbf{Pasolli E}, \textbf{Truong DT}, \textbf{Malik F},
\textbf{Waldron L}, \textbf{Segata N}. 2016. Machine learning
meta-analysis of large metagenomic datasets: Tools and biological
insights. PLoS Comput Biol \textbf{12}.
doi:\href{https://doi.org/10.1371/journal.pcbi.1004977}{10.1371/journal.pcbi.1004977}.

\hypertarget{ref-sze_looking_2016}{}
8. \textbf{Sze MA}, \textbf{Schloss PD}. 2016. Looking for a signal in
the noise: Revisiting obesity and the microbiome. mBio \textbf{7}.
doi:\href{https://doi.org/10.1128/mBio.01018-16}{10.1128/mBio.01018-16}.

\hypertarget{ref-walters_meta-analyses_2014}{}
9. \textbf{Walters WA}, \textbf{Xu Z}, \textbf{Knight R}. 2014.
Meta-analyses of human gut microbes associated with obesity and IBD.
FEBS Lett \textbf{588}:4223--4233.
doi:\href{https://doi.org/10.1016/j.febslet.2014.09.039}{10.1016/j.febslet.2014.09.039}.

\hypertarget{ref-vazquez-baeza_guiding_2018}{}
10. \textbf{Vázquez-Baeza Y}, \textbf{Gonzalez A}, \textbf{Xu ZZ},
\textbf{Washburne A}, \textbf{Herfarth HH}, \textbf{Sartor RB},
\textbf{Knight R}. 2018. Guiding longitudinal sampling in IBD cohorts.
Gut \textbf{67}:1743--1745.
doi:\href{https://doi.org/10.1136/gutjnl-2017-315352}{10.1136/gutjnl-2017-315352}.

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  11pt,
]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={A framework for effective application of machine learning to a microbiome-based classification problem},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{caption}
\usepackage{hyperref}
\usepackage{helvet} % Helvetica font
\renewcommand*\familydefault{\sfdefault} % Use the sans serif version of the font
\usepackage[T1]{fontenc}
\usepackage[labelfont=bf]{caption}

@@

\title{\textbf{A framework for effective application of machine learning to a
microbiome-based classification problem}}
\author{}
\date{\vspace{-2.5em}}

@@

\newpage
\linenumbers

\hypertarget{abstract}{%
\subsection{Abstract}\label{abstract}}

@@

\hypertarget{importance}{%
\subsection{Importance}\label{importance}}

@@

\hypertarget{background}{%
\subsection{Background}\label{background}}

As the number of people represented in human microbiome datasets grow,
there is an increasing desire to use microbiome data to diagnose
disease. However, the structure of the human microbiome is remarkably
variable among individuals to the point where it is often difficult to
identify the bacterial populations that are associated with diseases
using traditional statistical models. For example it is not possible to
classify individuals as having healthy colons or colorectal lesions
using Bray-Curtis distances based on the 16S rRNA gene sequences
collected from fecal samples {[}Figure S1{]}. This variation is likely
due to the ability of many bacterial populations to fill the same niche
such that different populations cause the same disease in different
individuals. Furthermore, a growing number of studies have shown that it
is rare for a single bacterial species to be associated with a disease.
Instead, subsets of the microbiome account for differences in health.
Traditional statistical approaches do not adequately account for the
variation in the human microbiome and typically consider the protective
or risk effects of each bacterial population separately (1). Recently,
machine learning (ML) models have grown in popularity among microbiome
researchers as our ability to sample large numbers of individuals has
grown; such models can effectively account for the interpersonal
microbiome variation and the ecology of disease.

@@

In this study, we provided steps toward standardization of machine
learning methods for microbiome studies which are often poorly
documented and executed. To showcase a rigorous ML pipeline and to shed
light on how ML model selection can affect modeling results, we
performed an empirical analysis comparing the predictive performance,
interpretability and training times of seven modeling approaches with
the same dataset and pipeline. We built three linear models with
different forms of regularization: L2-regularized logistic regression
and L1 and L2-regularized support vector machines (SVM) with a linear
kernel. We also trained four non-linear models: SVM with radial basis
function kernel, a decision tree, random forest, and gradient boosted
trees. We compared their predictive performance, interpretability, and
training time. To demonstrate the performance of these modeling
approaches and our pipeline on a case study, we used a data from a
previously published study that sought to classify individuals as having
healthy colons or colonic lesions based on the 16S rRNA gene sequences
collected from fecal samples (4). This dataset was selected because it
is a relatively large collection of individuals (N=490) connected to a
clinically significant disease where there is ample evidence that the
disease is driven by variation in the microbiome (2, 4, 5, 32). With
this dataset, we developed an ML pipeline that can be used in many
different scenarios for training and evaluating models. This framework
can be easily applied to other host-associated and environmental
microbiome datasets. We also provided an aspirational rubric for
evaluating the rigor of ML practices applied to microbiome data {[}Table
S1{]} to urge the audience to be diligent with with their study design,
model selection, development, evaluation and interpretation.

\hypertarget{results}{%
\subsection{Results}\label{results}}

@@

Model selection requires tuning hyperparameters. Hyperparameters are
parameters that need to be specified or tuned by the user, in order to
train a model for a specific modeling problem. For example, when using
regularization, C is a hyperparameter that indicates the penalty for
overfitting. Hyperparameters are tuned using the training data to find
the best model. We selected hyperparameters by performing repeated
five-fold cross-validation (CV) on the training set {[}Figure 1{]}. The
five-fold CV was also stratified to maintain the overall case and
control distribution. We chose the hyperparameter values that led to the
best average CV predictive performance using the area under the receiver
operating characteristic curve (AUROC) {[}Figure S2 and S3{]}. The AUROC
ranges from 0, where the model's predictions are perfectly incorrect, to
1.0, where the model perfectly distinguishes between cases and controls.
An AUROC value of 0.5 indicates that the model's predictions are no
different than random. To select hyperparameters, we performed a grid
search for hyperparameter settings when training the models. Default
hyperparameter settings in developed ML packages available in R, Python,
and MATLAB programming languages may be inadequate for effective
application of classification algorithms and need to be optimized for
each new ML task. For example, L1-regularized SVM with linear kernel
showed large variability between different regularization strengths (C)
and benefited from tuning as the default C parameter was 1 {[}Figure
S2{]}.

@@

\textbf{Predictive performance and generalizability of the seven
models.} We evaluated the predictive performance of the seven models to
classify individuals as having healthy colons or SRNs {[}Figure 2{]}.
The predictive performance of random forest model was higher than other
ML models with a median 0.695 {[}IQR 0.650-0.739{]}, though not
significantly (p=0.5; The p-value was manually calculated using the
sampling distribution of the test statistic under the null hypothesis)
(Figure S4). Similarly, L2-regularized logistic regression, XGBoost,
L2-regularized SVM with linear and radial basis function kernel AUROC
values were not significantly different from one another and had median
AUROC values of 0.680 {[}IQR 0.639-0.750{]}, 0.679 {[}IQR
0.643-0.746{]}, 0.678 {[}IQR 0.639-0.750{]} and 0.668 {[}IQR
0.639-0.750{]}, respectively. L1-regularized SVM with linear kernel and
decision tree had significantly lower AUROC values than the other ML
models with median AUROC of 0.650 {[}IQR 0.629-0.760{]} and 0.601 {[}IQR
0.636-0.753{]}, respectively {[}Figure 2{]}. Interestingly, these
results demonstrate that the most complex model (XGBoost) did not have
the best performance and that the most interpretable models
(L2-regularized logistic regression and L2-regularized SVM with linear
kernel) performed nearly as well as non-linear models.

@@

To show the effect of sample size on model generalizability, we compared
cross-validation AUROC values of L2-regularized logistic regression and
random forest models when we subsetted our original study design with
490 subjects to 15, 30, 60, 120, and 245 subjects {[}Figure S5{]}. The
variation in cross-validation performance within both models at smaller
sample sizes was larger than when the full collection of samples was
used to train and validate the models. Because of the high
dimensionality of the microbiome data (6920 OTUs), large sample sizes
can lead to better models.

\textbf{Interpretation of each ML model.} We often use ML models not
just to predict a health outcome, but also to identify potential
biomarkers for disease. Therefore, model interpretation becomes crucial
for microbiome studies. Interpretability is related to the degree to
which humans can understand the reasons behind a model prediction
(33--35). ML models often decrease in interpretability as they increase
in complexity. In this study, we used two methods to help interpret our
models.

@@

To further highlight the differences between the two interpretation
methods, we used permutation importance to interpret the linear models
{[}Figure S7{]}. When we analyzed the L1-regularized SVM with linear
kernel model using feature rankings based on weights {[}Figure 3{]} and
permutation importance {[}Figure S7{]}, 17 of the 20 top OTUs (e.g., OTU
609, 822, 1239) were deemed important by both interpretation methods.
Similarly, for the L2-regularized SVM and L2-regularized logistic
regression, 9 and 12 OTUs, respectively, were shared among the two
interpretation methods. These results indicate that both methods are
consistent in selecting the most important OTUs.

@@

\hypertarget{discussion}{%
\subsection{Discussion}\label{discussion}}

@@

\hypertarget{materials-and-methods}{%
\subsection{Materials and Methods}\label{materials-and-methods}}

@@

\textbf{16S rRNA gene sequencing data.} Stool samples provided by the
patients were used for 16S rRNA gene sequencing to measure bacterial
population abundances. The sequence data used in our analyses were
originally generated by Baxter et al.~(available through NCBI Sequence
Read Archive {[}SRP062005{]}, (5)). The OTU abundance table was
generated by Sze et al (45), who processed the 16S rRNA sequences in
mothur (v1.39.3) using the default quality filtering methods,
identifying and removing chimeric sequences using VSEARCH, and assigning
to OTUs at 97\% similarity using the OptiClust algorithm (43, 46, 47);
(\url{https://github.com/SchlossLab/Sze_CRCMetaAnalysis_mBio_2018/blob/master/data/process/baxter/baxter.0.03.subsample.shared}).
These OTU abundances were the features we used to predict colorectal
health of the patients. There were 6920 OTUs. OTU abundances were
subsampled to the size of the smallest sample and normalized across
samples such that the highest abundance of each OTU would be 1, and the
lowest would be 0.

@@

\textbf{Statistical analysis workflow.} Data summaries, statistical
analysis, and data visualizations were performed using R (v.3.5.0) with
the tidyverse package (v.1.2.1). We compared the performance of the
models pairwise by calculating the difference between AUROC values from
the same data-split (for 100 data-splits). We determined if the models
were significantly different by calculating the empirical p-value (2 x
min(\% of AUROC differences \(\geq\) 0, \% of AUROC differences \(\leq\)
0) for the double tail event (e.g., Figure S4).

@@

\hypertarget{references}{%
\subsection{References}\label{references}}

\hypertarget{refs}{}
\leavevmode\hypertarget{ref-segata_metagenomic_2011}{}%
1. \textbf{Segata N}, \textbf{Izard J}, \textbf{Waldron L},
\textbf{Gevers D}, \textbf{Miropolsky L}, \textbf{Garrett WS},
\textbf{Huttenhower C}. 2011. Metagenomic biomarker discovery and
explanation. Genome Biol \textbf{12}:R60.
doi:\href{https://doi.org/10.1186/gb-2011-12-6-r60}{10.1186/gb-2011-12-6-r60}.

\leavevmode\hypertarget{ref-zeller_potential_2014}{}%
2. \textbf{Zeller G}, \textbf{Tap J}, \textbf{Voigt AY},
\textbf{Sunagawa S}, \textbf{Kultima JR}, \textbf{Costea PI},
\textbf{Amiot A}, \textbf{Böhm J}, \textbf{Brunetti F},
\textbf{Habermann N}, \textbf{Hercog R}, \textbf{Koch M},
\textbf{Luciani A}, \textbf{Mende DR}, \textbf{Schneider MA},
\textbf{Schrotz-King P}, \textbf{Tournigand C}, \textbf{Tran Van Nhieu
J}, \textbf{Yamada T}, \textbf{Zimmermann J}, \textbf{Benes V},
\textbf{Kloor M}, \textbf{Ulrich CM}, \textbf{Knebel Doeberitz M von},
\textbf{Sobhani I}, \textbf{Bork P}. 2014. Potential of fecal microbiota
for early-stage detection of colorectal cancer. Mol Syst Biol
\textbf{10}.
doi:\href{https://doi.org/10.15252/msb.20145645}{10.15252/msb.20145645}.

\leavevmode\hypertarget{ref-zackular_human_2014}{}%
3. \textbf{Zackular JP}, \textbf{Rogers MAM}, \textbf{Ruffin MT},
\textbf{Schloss PD}. 2014. The human gut microbiome as a screening tool
for colorectal cancer. Cancer Prev Res \textbf{7}:1112--1121.
doi:\href{https://doi.org/10.1158/1940-6207.CAPR-14-0129}{10.1158/1940-6207.CAPR-14-0129}.

\leavevmode\hypertarget{ref-baxter_dna_2016}{}%
4. \textbf{Baxter NT}, \textbf{Koumpouras CC}, \textbf{Rogers MAM},
\textbf{Ruffin MT}, \textbf{Schloss PD}. 2016. DNA from fecal
immunochemical test can replace stool for detection of colonic lesions
using a microbiota-based model. Microbiome \textbf{4}.
doi:\href{https://doi.org/10.1186/s40168-016-0205-y}{10.1186/s40168-016-0205-y}.

\leavevmode\hypertarget{ref-baxter_microbiota-based_2016}{}%
5. \textbf{Baxter NT}, \textbf{Ruffin MT}, \textbf{Rogers MAM},
\textbf{Schloss PD}. 2016. Microbiota-based model improves the
sensitivity of fecal immunochemical test for detecting colonic lesions.
Genome Medicine \textbf{8}:37.
doi:\href{https://doi.org/10.1186/s13073-016-0290-3}{10.1186/s13073-016-0290-3}.

\leavevmode\hypertarget{ref-hale_shifts_2017}{}%
6. \textbf{Hale VL}, \textbf{Chen J}, \textbf{Johnson S},
\textbf{Harrington SC}, \textbf{Yab TC}, \textbf{Smyrk TC},
\textbf{Nelson H}, \textbf{Boardman LA}, \textbf{Druliner BR},
\textbf{Levin TR}, \textbf{Rex DK}, \textbf{Ahnen DJ}, \textbf{Lance P},
\textbf{Ahlquist DA}, \textbf{Chia N}. 2017. Shifts in the fecal
microbiota associated with adenomatous polyps. Cancer Epidemiol
Biomarkers Prev \textbf{26}:85--94.
doi:\href{https://doi.org/10.1158/1055-9965.EPI-16-0337}{10.1158/1055-9965.EPI-16-0337}.

\leavevmode\hypertarget{ref-pasolli_machine_2016}{}%
7. \textbf{Pasolli E}, \textbf{Truong DT}, \textbf{Malik F},
\textbf{Waldron L}, \textbf{Segata N}. 2016. Machine learning
meta-analysis of large metagenomic datasets: Tools and biological
insights. PLoS Comput Biol \textbf{12}.
doi:\href{https://doi.org/10.1371/journal.pcbi.1004977}{10.1371/journal.pcbi.1004977}.

\leavevmode\hypertarget{ref-sze_looking_2016}{}%
8. \textbf{Sze MA}, \textbf{Schloss PD}. 2016. Looking for a signal in
the noise: Revisiting obesity and the microbiome. mBio \textbf{7}.
doi:\href{https://doi.org/10.1128/mBio.01018-16}{10.1128/mBio.01018-16}.

\leavevmode\hypertarget{ref-walters_meta-analyses_2014}{}%
9. \textbf{Walters WA}, \textbf{Xu Z}, \textbf{Knight R}. 2014.
Meta-analyses of human gut microbes associated with obesity and IBD.
FEBS Lett \textbf{588}:4223--4233.
doi:\href{https://doi.org/10.1016/j.febslet.2014.09.039}{10.1016/j.febslet.2014.09.039}.

\leavevmode\hypertarget{ref-vazquez-baeza_guiding_2018}{}%
10. \textbf{Vázquez-Baeza Y}, \textbf{Gonzalez A}, \textbf{Xu ZZ},
\textbf{Washburne A}, \textbf{Herfarth HH}, \textbf{Sartor RB},
\textbf{Knight R}. 2018. Guiding longitudinal sampling in IBD cohorts.
Gut \textbf{67}:1743--1745.
doi:\href{https://doi.org/10.1136/gutjnl-2017-315352}{10.1136/gutjnl-2017-315352}.
